\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-04-18 docs/papers/sto/so.tex
Update paper draft

% Main text: up to 3500 words: introduction + results + discussion
% so we should keep introduction and discussion brief
\vspace{1em}
\section{Introduction}
% without heading

@@

% combination of differentiable simulations, NN, and SR
NNs are flexible and efficient universal function approximators
\citep{Hornik1989}, which can be used to learn physical processes
when embedded in numerical simulations.
NNs are typically trained with gradient based optimizers, which requires
the simulations to be differentiable.

@@

% introduce PM and its limitation on small scales
The particle-mesh (PM) method is a fundamental technique for
cosmological $N$-body simulations.
PM solves the Poisson equation by scattering the particles onto an
auxiliary mesh, on which the fast Fourier transform (FFT) is performed
and the Laplacian operator reduces to a simple multiplication of
wavenumbers in Fourier space.
The cloud-in-cell (CIC) scheme \citep{HockneyEastwood1988} has been
widely used to compute the mass fractions of a particle at $\vx'$
scattered to a nearby mesh grid point at $\vx$
%
\begin{equation}
  W(\vx, \vx') = \prod_{i=1}^3
    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
  \label{cic}
  \end{equation}
%
where $l_c$ is the mesh cell size.
The gravitational forces on the mesh points are then grathered to form
the interpolated force for each particle, with the same weights as in
the scattering process.
Other scattering schemes include the nearest grid point (NGP) and
triangle shaped cloud (TSC) etc.
Although being computationally efficient, the PM method is limited by
its resolution on small scales, i.e. scales around and below the mesh
cell size.
This is due to the loss of information during the scattering and
gathering process between the continuously distributed particles and the
discrete mesh.

% motivation of this work
Optimizing the accuracy of fast and differentiable simulations is
crucial for their applications in, e.g. generation of mocks for
cosmological surveys, simulation-based inference or field-level
inference.

% introduction of this work
The differentiable cosmological simulation with adjoint method
\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
\citep{Li2022b} lay the foundation of the gradient-based optimization
discussed in this work.
Within this fully differentiable framework, we sharpen the gravitational
force by embedding NNs in the Poisson equation and train them for the
spatial optimization (SO) of the simulation resolution on small scales.
We further interpret the trained NNs using symbolic regression (SR).

@@

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational force.
The dynamics of the particles is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $\vF = - \vnabla \varphi$
follows
%
\begin{equation}
\vF(\vk) = - i \vk \varphi(\vk) =
                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\label{force}
\end{equation}
%
which then gives the force field in configuration space after an inverse
Fourier transform.

In the PM simulation, the density field $\delta(\vx)$ is estimated by
scattering the particles onto the mesh with e.g. the CIC scheme in
\eqref{cic}.
Then the force on each particle is estimated by gathering the force
field from the mesh.
Therefore, the accuracy of the gravitational force estimated for each
particle depends on the resolution of the mesh.
To improve the accuracy of the PM solver, we introduce SO force
sharpening by modifying the force equation \eqref{force} to
%
\begin{equation}
F(\vk)_i = i \frac32 \Omegam H_0^2 \,\frac{k_i}{k^2}\,
                    f(k_i; \vvartheta) g(\vk; \vvartheta) \delta(\vk),
\end{equation}
%
where $f$ and $g$ are nonlinear functions of the wavevector components
and all possible relevant features $\vvartheta$, which are approximated
with NNs.
$k_i$ is a component of $\vk$, with $i\in\{1,2,3\}$ denoting the 3
principal axes of the simulation box.

>>>>

% Main text: up to 3500 words: introduction + results + discussion
\vspace{1em}
\section{Introduction}
% without heading

@@

@@

Cosmological $N$-body simulations evolve dark matter particles under the
gravitational interaction.
The dynamics of the dark matter is governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $\vF = - \vnabla
\varphi$ follows
%
\begin{equation}
\vF(\vk) = - i \vk \varphi(\vk) =
                  i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\label{force}
\end{equation}
%
where the Laplacian operator reduces to simple multiplications.
The force field in configuration space $\vF(\vx)$ can then be recovered
with an inverse Fourier transform.
The particle-mesh (PM) method is a fundamental and efficient approach to
solving the Poisson equation given numerous particles.
In order to use the fast Fourier transform (FFT), PM estimates the
matter density field on a regular spatial grid by scattering the
particles onto an auxiliary mesh.
The cloud-in-cell (CIC), or trilinear, interpolation scheme
\citep{HockneyEastwood1988} has been widely used to compute the mass
fractions of a particle at $\vx'$ scattered to a nearby mesh point
at $\vx$
%
\begin{equation}
  W(\vx, \vx') = \prod_{i=1}^3
    \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_c}, 0 \Bigr),
  \label{cic}
  \end{equation}
%
where $l_c$ is the mesh cell size.
Other scattering schemes include the nearest grid point (NGP) and
triangle shaped cloud (TSC) etc.
The gravitational forces on the mesh points are computed in the Fourier
space and transformed back to the configuration space, which are then
grathered to form the interpolated force for each particle, with the
same weights as in the scattering process.
Although being computationally efficient, the accuracy of the
gravitational force estimated for each particle depends on the
resolution of the mesh.
The PM method is mainly limited by its resolution on small scales, i.e.
scales around and below the mesh cell size.
This is due to the loss of information during the scattering and
gathering process between the continuously distributed particles and the
regular discrete mesh.

In this work, we sharpen the gravitational force by embedding NNs in the
force equation and train them for the spatial optimization (SO) of the
PM simulation.
NNs are flexible and efficient universal function approximators
\citep{Hornik1989}, which can be used to learn physical processes
when embedded in numerical simulations.
NNs are typically trained with gradient based optimizers, which requires
the simulations to be differentiable.
The differentiable cosmological simulation with adjoint method
\citep{Li2022a} and the corresponding PM $N$-body library \pmwd\
\citep{Li2022b} lay the foundation of the gradient-based optimization
discussed in this work.
We further interpret the trained NNs using symbolic regression (SR).
Optimizing the accuracy of fast and differentiable simulations is
crucial for their applications in, e.g. generation of mocks for
cosmological surveys, simulation-based inference or field-level
inference.

@@

To improve the accuracy of the PM solver, we introduce SO force
sharpening by modifying the gravitational force equation \eqref{force}
to
%
\begin{equation}
F(\vk)_i = i \frac32 \Omegam H_0^2 \,\frac{k_i}{k^2}\,
                    f(k_i; \vvartheta) g(\vk; \vvartheta) \delta(\vk),
\end{equation}
%
where $f$ and $g$ are nonlinear functions of the wavevector components
and all possible relevant features $\vvartheta$, which are approximated
with NNs.
$k_i$ is a component of $\vk$, with $i\in\{1,2,3\}$ denoting the 3
principal axes of the simulation box.
